D3064R0: OOTA Execution is Provably Vacuous

### 2023-12-06 ootavacuous.tex
Relocate the Illustrative Examples section
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{listing}[tbp]
\begin{verbatim}
 1 int main() {
 2   atomic_int x=0;
 3   atomic_int y=0;
 4   int r1;
 5   int r2;
 6   {{{ { r1=x.load(memory_order_relaxed);
 7         y.store(r1,memory_order_relaxed); }
 8   ||| { r2=y.load(memory_order_relaxed);
 9         x.store(r2,memory_order_relaxed); }  }}}
10   return 0;
11 }
\end{verbatim}
\caption{OOTA Cycle}
\label{lst:OOTA Cycle}
\end{listing}

Listing~\ref{lst:OOTA Cycle}
shows an OOTA cycle, where the \co{\{\{\{} on line~5 begins a set of
threads, the \co{|||} on line~8 separates a pair of threads, and
the \co{\}\}\}} on line~9 ends a set of threads.\footnote{
	This is pseudocode in the format of CPPMEM,
	which does not permit writes from non-local variables.}
Lines~6 and~7 do a relaxed copy from \co{x} to \co{y}, and
lines~8 and~9 to a relaxed copy from \co{y} to \co{x}.
Even though both \co{x} and \co{y} are initialized to zero on
lines~2 and~3, respectively, according to the mathematical core
of the C++ memory model, the only constraint on the final values
of \co{x} and \co{y} is that they be equal.
Counterintuitivily, their final values could be any member of their type.

\begin{listing}[tbp]
\begin{verbatim}
 1 int main() {
 2   atomic_int x = 0;
 3   atomic_int y = 0;
 4   volatile int r1;
 5   volatile int r2;
 6   {{{ { r1 = x.load(memory_order_relaxed);
 7         if (r1 == 1)
 8           y.store(1, memory_order_relaxed);
 9         else
10           y.store(r1, memory_order_relaxed); }
11   ||| { r2=y.load(memory_order_relaxed);
12         x.store(r2,memory_order_relaxed); }  }}}
13   }}};
14   return 0;
15 }
\end{verbatim}
\caption{OOTA Cycle Based on Conditionals}
\label{lst:OOTA Cycle Based on Conditionals}
\end{listing}

A similar effect can be achieved using conditional statements, as
shown in Listing~\ref{lst:OOTA Cycle Based on Conditionals}.

@@

C++ implementations may of course avail themselves of the as-if rule.
However, correct use of the as-if rule requires that the observable
behaviors of the program be appropriately preserved.

@@

We therefore define an OOTA cycle (also known as a causal cycle) as a
cycle in the memory-model sense involving only reads-from links, that
is, the value stored in on thread being read by the next thread.
Within each thread in the cycle, the value stored depends on
the value loaded.
This dependency might be a data dependency
(see Listing~\ref{lst:OOTA Cycle})
or it might be a control dependency
(see Listing~\ref{lst:OOTA Cycle Based on Conditionals}).\footnote{
	Sufficiently motivated readers should have no problem constructing
	examples involving address dependencies.}
@@@ There must be at least two distinct values such that if a given load
returns a given value, that value forces the cycle to execute so as to
cause the corresponding store's value to be equal to that given value, and
for that store to occur early enough for the load to return that value.

@@

@@

\subsection{Illustrative Examples}
\label{sec:Illustrative Examples}

This section works through OOTA examples, showing the effects of
the constraints called out in this paper.

\subsubsection{OOTA Cycle}
\label{sec:OOTA Cycle}

Listing~\ref{lst:OOTA Cycle}
shows the canonical example of an OOTA cycle.
Because values from speculative stores cannot be visible to other threads,
line~7 cannot be committed until after line~6 completes.
Similarly, line~9 cannot be committed until after line~8 completes.

Because values stored take non-zero time to propagate from one thread
to another, no cycle can form, and thus no OOTA cycle can form.

\subsubsection{OOTA Cycle Based on Conditionals}
\label{sec:OOTA Cycle Based on Conditionals}

Listing~\ref{lst:OOTA Cycle Based on Conditionals}
shows an if-condition-based example of an OOTA cycle.
Because values from speculative stores cannot be visible to other threads,
lines~8 and~10 cannot be committed until after the load in line~6 completes
and the conditional on line~7 is evaluated non-speculatively.
Similarly, line~12 cannot be committed until after line~11 completes.

Because values stored take non-zero time to propagate from one thread
to another, no cycle can form, and thus no OOTA cycle can form.

Note that if both legs of the \co{if} statement spanning lines~7-10,
the implementation would be within its rights to hoist the store out of
that \co{if} statement.
However, in this case, the two stores on lines~8 and~10 are storing
distinct values, preventing the hoisting.

>>>>

\begin{listing}[tbp]
@@ DisplayLitmus litmus/oota-ctrl.litmus @@
\caption{OOTA Cycle}
\label{lst:OOTA Cycle}
\end{listing}

Listing~\ref{lst:OOTA Cycle}
shows an OOTA cycle.
The first line identifies it as a C-language litmus test and gives it
a name.
Lines~2-5 initialize variables, in this case setting the initial
values of the global shared variables \co{x} and \co{y} to zero.
Lines~7-13 define the first process, \co{P0()}, and lines~15-21
define the second process, \co{P1()}.
The arguments to both \co{P0()} and \co{P1()} specify which of
the global shared variables each process may access.
In this case, both processes may access both \co{x} and \co{y}.
The body of each process specifies a limited subset of C++ code,
which also happens to be a somewhat less limited subset of C code.

Finally, line~23 specifies an \co{exists} clause, which gives a
condition to check for the final value of the specified local
variables.
The \co{0:} prefix specifies a variable local to \co{P0()} and
the \co{1:} prefix specifies a variable local to \co{P1()}.
The \co{/\\} specifies a boolean AND, and the \co{=} specifies
an equality comparison.

\begin{listing}[tbp]
@@ RunLitmus litmus/oota-ctrl.litmus @@
\caption{OOTA Cycle, \co{herd7} Output}
\label{lst:OOTA Cycle, herd7 Output}
\end{listing}

Line~4 of the corresponding output of the \co{herd7} tool shows the
counterintuitive outcome where both processes load the value 42.
As we will see, this outcome is not possible in real-world implementations
due to physical constraints of which \co{herd7} is (by design) unaware.

@@

@@

We therefore define an OOTA cycle (also known as a causal cycle) as a
cycle in the memory-model sense involving only reads-from links, that
is, the value stored in on thread being read by the next thread.
Within each thread in the cycle, the value stored depends on
the value loaded.
This dependency might be a control dependency
(see Listing~\ref{lst:OOTA Cycle})
or it might be a data dependency
(see Listing~\ref{lst:Causality Test Case 4} on
page~\pageref{lst:Causality Test Case 4}).\footnote{
	Sufficiently motivated readers should have no problem constructing
	examples involving address dependencies.}
@@@ There must be at least two distinct values such that if a given load
returns a given value, that value forces the cycle to execute so as to
cause the corresponding store's value to be equal to that given value, and
for that store to occur early enough for the load to return that value.

@@

Applying this to Listing~\ref{lst:OOTA Cycle}, because values from
speculative stores cannot be visible to non-speculative loads, lines~10
and~12 cannot be committed until after the load in line~8 completes
and the conditional on line~9 is evaluated non-speculatively.\footnote{
	Alternatively, one could imagine that both the speculation
	and the squashing of failed speculation might span multiple
	threads.}
Similarly, lines~18 and~20 cannot be committed until after lines~16
and~17 complete non-speculatively.

Note that if both legs of the \co{if} statement spanning lines~9-12 were
to store the same value, the implementation would be within its rights
to hoist the store out of that \co{if} statement.
However, in this case, the two stores on lines~10 and~12 are storing
distinct values, preventing the hoisting.

@@
